---
title: "Class 9: Candy Mini-Project"
author: "Nathalie Huang (PID: A19134713)"
format: pdf
toc: TRUE
---

## Importing candy data
```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` rows in this dataset

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

One of the most interesting variables in the dataset is winpercent. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset. Higher values indicate a more popular candy.

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Sour Patch Kids", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset.

#install skimr package
#install.packages("skimr")

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Most variables in the candy dataset are measured on a zero to one scale, either as binary indicators or proportions. However, the variable winpercent appears to be on a different scale compared to the majority of the dataset. Unlike the other columns, winpercent ranges from approximately 0 to 100 and represents the percentage of people who prefer each candy. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

The chocolate column is a binary variable that only takes values of zero or one. A value of one indicates that the candy contains chocolate, while a value of zero indicates that it does not. It tells us whether chocolate is an ingredient in the candy.

## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
hist(candy$winpercent,
     main = "Histogram of Candy Win Percentages",
     xlab = "Win Percent")
```

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(color="black", fill="lightgray") +
  labs(title = "Histogram of Candy Win Percentages",
       x = "Win Percent",
       y = "Count")
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values does not appear to be symmetrical. While the data are somewhat centered around the middle range, the histogram shows a slight skew. This suggests a mild right skew rather than a symmetric, bell-shaped distribution. This means that most candies have lower winpercent values, and only a smaller number of candies have very high winpercent values.

> Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50%.

Welch Two Sample t-test

```{r}
chocolate <- candy$winpercent[as.logical(candy$chocolate)]
fruity <- candy$winpercent[as.logical(candy$fruity)]

t.test(chocolate, fruity)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Steps to solve this:
1. Find all chocolate candy in the dataset
2. Extract or find their winpercent values
3. Calculate the mean of these values
4. Find all fruit candy
5. Find their winpercent values
6. Calculate their mean value

```{r}
choc.candy <- candy[candy$chocolate==1, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```
```{r}
fruity.candy <- candy[candy$fruity==1, ]
fruity.win <- fruity.candy$winpercent
mean(fruity.win)
```

The mean winpercent for chocolate candies is approximately 60.9%, whereas the mean for fruity candies is about 44.1%. This indicates that chocolate candies are generally more preferred than fruity candies in this dataset.

> Q12. Is this difference statistically significant?

Yes, the difference is statistically significant. The Welch two-sample t-test comparing chocolate and fruity candies produced a p-value below the conventional significance threshold of 0.05. Additionally, the 95% confidence interval for the difference in means does not include zero, providing evidence that chocolate candies are significantly more popular than fruity candies on average.

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent), ], n = 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent), ], n = 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x = winpercent, y = pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 100)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Based on the plot of winpercent versus pricepercent, Reese’s Miniatures appear to offer the most bang for your buck. They have one of the highest winpercent values while also having a relatively low pricepercent.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The five most expensive candy types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Among these higher-priced candies, Nik L Nip is the least popular, as it has the lowest winpercent of the group.

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are the most anti‑correlated pair. They are the deepest red circle (towards -1).

> Q23. Similarly, what two variables are most positively correlated?

The variables chocolate and bar show a strong positive correlation, since many chocolate candies are sold in bar form. In addition, chocolate and winpercent are also positively correlated, as candies that contain chocolate tend to have higher win percentages and are generally more popular.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)

plot(pca$x[, 1:2])
```

#We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

# Make a new data-frame with our PCA results and candy data
```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x = PC1, y = PC2, 
            size = winpercent / 100,  
            text = rownames(my_data),
            label = rownames(my_data)) +
        geom_point(col = my_cols)

p
```

Again we can use the ggrepel package and the function ggrepel::geom_text_repel() to label up the plot with non overlapping candy names like.

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 100)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

#install.packages("plotly")

#```{r}
#library(plotly)
#ggplotly(p)


```{r}
ggplot(pca$rotation) +
  aes(x = PC1, y = rownames(pca$rotation)) +
  geom_col()
```
```{r}
ggplot(pca$rotation) +
  aes(x = PC1, y = reorder(rownames(pca$rotation), PC1)) +
  geom_col() +
  labs(x = "PC1", y = "") +
  theme_minimal()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The variables most strongly associated with the positive direction of PC1 are fruity, pluribus, and hard. This makes sense because these features tend to co‑occur in non‑chocolate, sugar‑based candies, which often come as multiple small pieces and have a harder texture. PC1 is capturing a contrast between fruity candies and chocolate candies. We saw this same relationship earlier in the correlation matrix, where fruity and chocolate were strongly anti‑correlated, and chocolate and bar were positively correlated.

## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The candies that tend to “win” are those with classic chocolate bar characteristics. Chocolate candies consistently showed higher winpercent values than fruity candies, and the correlation analysis highlighted strong positive relationships between chocolate and other bar‑like traits such as caramel, nougat, and peanutyalmondy. These traits were negatively correlated with fruity and hard‑candy features, suggesting a clear divide between the two candy styles. The PCA results reinforced this pattern by placing chocolate‑bar attributes together on one side of PC1—along with winpercent—while fruity, pluribus, and hard candies loaded strongly in the opposite direction.


